---
title: "Gov 50 Final Project"
author: "Molly Malague"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



```{r, echo = FALSE}
library(tidyverse)
```

## Research Question/Thoughts!!!

What effect does a lack of health insurance have on a community's average life expectancy?
In this study, I plan to examine the correlation between health insurance coverage and life expectancy, across counties in the New England states. I hypothesize that communities with less health insurance coverage will have lower average life expectancies. 
The explanatory variable of interest is statistics on forms/extent of life expectancy. My outcome variable is the life expectancy in the same area, as measured by the US Dept. of Health and Human Services and Census data. 
If I observe that greater health insurance coverage aligns with longer average life expectancies, this would provide support for my hypothesis. If, conversely, I observe no correlation between health insurance and life expectancy, this would provide evidence against my hypothesis. 


```{r, results='hide', echo = FALSE}
coverage <- read_csv("Health_Insurance_Coverage_-_Counties_2015-2019.csv") 
###
 good_names <- coverage |>
   rename("Employer Insurance only - Pop 35-64" = "B27010_036E",
          "Employer Insurance only - Pop 19-34" = "B27010_020E",
          "Employer Insurance only - Pop 65+" = "B27010_053E",
          "Medicare only - Pop 19-34" = "B27010_022E",
          "Medicare only - Pop 35-64" = "B27010_038E",
          "Medicare only - Pop 65+" = "B27010_055E",
          "Medicaid only - Pop 19-34" = "B27010_023E",
          "Medicaid only - Pop 35-64" = "B27010_039E",
          "Medicaid only - Pop 65+" = "MEDICAID65_CALC",
          "Population with no Health Insurance Ages - 19-34" = "B27010_033E",
          "Population with no Health Insurance Ages - 35-64" = "B27010_050E",
          "Population with no Health Insurance Ages - Age 65+" = "B27010_066E",
          "Percent of Population with No Health Insurance Coverage" = "DP03_0099PE")
### 
 good_names_only <- good_names |>
  select("STATEFP", "Employer Insurance only - Pop 35-64", "Employer Insurance only - Pop 19-34", "Employer Insurance only - Pop 65+", "Medicare only - Pop 19-34", "Medicare only - Pop 35-64", "Medicare only - Pop 65+",  "Medicaid only - Pop 19-34",  "Medicaid only - Pop 35-64", "Medicaid only - Pop 65+", "Population with no Health Insurance Ages - 19-34", "Population with no Health Insurance Ages - 35-64", "Population with no Health Insurance Ages - Age 65+", "Percent of Population with No Health Insurance Coverage", "COUNTYFP", "COUNTYNS", "AFFGEOID", "GEOID", "NAME", "LSAD", "ALAND", "AWATER", "GEO_PARENT_NAME", "POP_DENSITY")
###
new_england <- c(23, 09, 25, 33, 44, 50)
###
me_LI <- good_names_only  |>
  filter(STATEFP %in% new_england)
me_LI
###
BESTLE <- read_csv("U.S._Life_Expectancy_at_Birth_by_State_and_Census_Tract_-_2010-2015.csv")
###
NE_names <- c("Maine", "New Hampshire", "Connecticut", "Vermont", "Massachusetts", "Rhode Island")
###
yay <- BESTLE |>
  filter(State %in% NE_names)
yay
###
library(dplyr)
me_LE <- yay |>
  drop_na(`Life Expectancy`) |>
  filter(State == NE_names) |>
  group_by(County) |> 
  summarise(mean_LE = mean(`Life Expectancy`))
me_LE
###
me_LE[c('County', 'State')] <- str_split_fixed(me_LE$County, ' ', 2)
me_LE
###
perfect <- me_LI |>
  inner_join(me_LE, by = c("NAME" = "County")) |>
  select(-c("LSAD", "State", "AWATER", "ALAND", "GEOID", "AFFGEOID"))
perfect
###
PERFECT <- perfect |>
  select(c("NAME", "STATEFP", "POP_DENSITY", "mean_LE",
               "Percent of Population with No Health Insurance Coverage", "Population with no Health Insurance Ages - Age 65+", "Population with no Health Insurance Ages - 35-64", "Population with no Health Insurance Ages - 19-34", "Medicaid only - Pop 35-64", "Medicaid only - Pop 19-34", "Medicare only - Pop 65+", "Medicare only - Pop 35-64", "Medicare only - Pop 19-34", "Employer Insurance only - Pop 65+", "Employer Insurance only - Pop 35-64", "Employer Insurance only - Pop 19-34"))
PERFECT
###

```

```{r, echo = FALSE}
PERFECT |>
  ggplot(aes(x = `Percent of Population with No Health Insurance Coverage`, y = mean_LE, color = STATEFP)) +
  geom_point() + geom_hline(aes(yintercept = mean(mean_LE)), color = "black", linetype = 3) + 
  geom_vline(aes(xintercept = mean(`Percent of Population with No Health Insurance Coverage`)), color = "black", linetype = 3) + 
  labs(x = "Percent of Population with No Health Insurance Coverage",
       y = "Mean Life Expectancy",
       title = "Percent of Population without Health Insurance -- and Average Life Expectancy",
       subtitle = "Among Counties in the Northeast")

```

```{r, results='hide', echo = FALSE}
summary(lm(mean_LE ~ `Percent of Population with No Health Insurance Coverage` , data = PERFECT))
```


```{r, echo = FALSE}
PERFECT |>
  ggplot(aes(x = `Percent of Population with No Health Insurance Coverage`, y = mean_LE)) +
   geom_smooth(method = "lm", se = FALSE) + geom_hline(aes(yintercept = mean(mean_LE)), color = "black", linetype = 3) + geom_vline(aes(xintercept = mean(`Percent of Population with No Health Insurance Coverage`)), color = "black", linetype = 3) +
  labs(x = "Percent of Population with No Health Insurance Coverage",
       y = "Mean Life Expectancy",
       title = "Percent of Population without Health Insurance -- and Average Life Expectancy",
       subtitle = "Among Counties in the Northeast")

```


```{r, echo = FALSE}
PERFECT |>
  ggplot(aes(x = `Percent of Population with No Health Insurance Coverage`, y = mean_LE, color = STATEFP)) +
   geom_smooth(method = "lm", se = FALSE) + geom_hline(aes(yintercept = mean(mean_LE)), color = "black", linetype = 3) + geom_vline(aes(xintercept = mean(`Percent of Population with No Health Insurance Coverage`)), color = "black", linetype = 3) + 
  labs(x = "Percent of Population with No Health Insurance Coverage",
       y = "Mean Life Expectancy",
       title = "Percent of Population without Health Insurance -- and Average Life Expectancy",
       subtitle = "Among Counties in the Northeast",
       color = "States") + 
scale_color_discrete(name = "State", labels = c("Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont")) 

```


discussion : I do not observe any meaningful correlation between health insurance overage and average life expectancy. There are a few patterns in individual states, but nothing clear across the regions. The region is large enough that it is likely representative of the nation writ large, so I am surprised to find my hypothesis refuted. 
a limitation in the scope of this analysis is the dataset's use of population totals, not percentages in 


## summary(lm(mean_LE ~ `Percent of Population with No Health Insurance Coverage`+, data = PERFECT))

## https://disasters.census.gov/datasets/USCensus::health-insurance-coverage-counties-2015-2019/about

##https://www.cdc.gov/nchs/data-visualization/life-expectancy/index.html 
